---
title: "Getting and Cleaning Data Using dplyr, tidyr"
author: "Arin Parsa"
date: "6/8/2020"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Manipulating Tabular Data with dplyr

Following exercises are from swirl() assignments by JHU. 
```{r}
install.packages("dplyr", repos='http://cran.us.r-project.org')
library(dplyr)
packageVersion("dplyr") #should be more than 1.0.0. If not install and load again

mydf <- read.csv("data/sample.csv", stringsAsFactors = FALSE)
head(library(dplyr))

cran <- as_tibble(mydf)
## "The main advantage to using a tbl_df over a regular data frame is the
## printing."
cran #output is more informative and compact

```
### select()
```{r}
select(cran, ip_id, package, country)
select(cran,r_arch:country)
select(cran,country:r_arch)
select(cran,-time)
select(cran, -(X:size))

```
### filter()
```{r}
filter(cran, package == "swirl")
filter(cran, r_version == "3.1.1", country == "US")
filter(cran, r_version <= "3.0.2", country == "IN")
filter(cran, country == "US" | country == "IN")
filter(cran, size > 100500, r_os == "linux-gnu")

```
### filtering out missing values
```{r}
is.na(c(3,5,NA,10))
!is.na(c(3,5,NA,10))
filter(cran,!is.na(r_version))

```
### arrange() for sorting
```{r}
cran2 <- select(cran,size:ip_id)
arrange(cran2, ip_id)
arrange(cran2, desc(ip_id))
arrange(cran2, package, desc(ip_id))
arrange(cran2, package, ip_id)
arrange(cran2, country, desc(r_version), desc(ip_id))
arrange(cran2, country, desc(r_version), ip_id)

```
### mutate() to add or remove columns based on an existing column
```{r}
cran3 <- select(cran, ip_id, package, size)
cran3
mutate(cran3, size_mb = size / 2^20)
mutate(cran3, size_mb = size / 2^20, size_gb = size_mb / 2^10)
mutate(cran3, correct_size = size + 1000)

```
### summarize() to calculate a new variable based on FOR EACH of the values in a column
```{r}
avg_bytes <- summary (cran, avg_bytes = mean(size))
summarize (cran, avg_bytes = mean(size))

```
## Data Grouping and Chaining with dplyr

### Grouping
```{r}
by_package <- group_by(cran, package)
by_package
summarize(by_package, mean(size))

pack_sum <- summarize(by_package,
                      count = n() ,
                      unique = n_distinct(ip_id),
                      countries = n_distinct(country),
                      avg_bytes = mean(size))

pack_sum
quantile(pack_sum$count, probs = 0.99)
View(cran)
top_counts <- filter(pack_sum, pack_sum$count > 679)
top_counts <- filter(pack_sum, count > 679)
top_counts
View(top_counts)
arrange(top_counts, desc(count))
top_counts_sorted <- arrange(top_counts, desc(count))
View(top_counts_sorted)
quantile(pack_sum$unique, probs = 0.99)
top_unique <- filter(pack_sum, unique > 465)
View(top_unique)
arrange(top_unique, desc(unique))
top_unique_sorted <- arrange(top_unique, desc(unique))



```
### Chaining
```{r}
result3 <-
  cran %>%
  group_by(package) %>%
  summarize(count = n(),
            unique = n_distinct(ip_id),
            countries = n_distinct(country),
            avg_bytes = mean(size)) %>%
  filter(countries > 60) %>%
  arrange(desc(countries), avg_bytes)
result3
```
## Tidying Data
What is meant by tidy data: http://vita.had.co.nz/papers/tidy-data.pdf

According to Hadley Wickam, tidy data is a standard way of mapping the meaning of a dataset to its structure. A dataset is messy or tidy depending on how rows, columns and tables are matched up with observations, variables and types. In tidy data:
1. Each variable forms a column.
2. Each observation forms a row.
3. Each type of observational unit forms a table.

### Five most common problems of messy data
* Column headers are values, not variable names.
* Multiple variables are stored in one column.
* Variables are stored in both rows and columns.
* Multiple types of observational units are stored in the same table.
* A single observational unit is stored in multiple tables.


```{r}
install.packages("readr", repos='http://cran.us.r-project.org')
install.packages("tidyr", repos='http://cran.us.r-project.org')
install.packages("dplyr", repos='http://cran.us.r-project.org')
library(readr)
library(tidyr)
library(dplyr)

```

###  Column headers are values, not variable names.
```{r}
students <- read.csv("data/students.csv") #file from swirl() exercise is in this repo 
students
gather(students, sex, count, -grade,-X)

```

### Multiple variables are stored in one column
```{r}
students2 <- read.csv("students2.csv")
students2
res <- gather(students2, sex_class, count, -grade)
res
separate(res, sex_class, c("sex", "class"))


```
### Variables are stored in both rows and columns.
```{r}

students3 <- read.csv("students3.csv")
students3
##class_1 through 5 columns will now be under 1 variable class
res <- gather(students3, class, grade, class1:class5, na.rm = TRUE)
res
## midterm and final grade rows will be in different variable columns (midterm, final with values being the grades)
res <-spread(res, test,grade )
res
#class1, class2...class5 values be stripped of "class" and just be the number
mutate(res, class = parse_number(class))


```
### A single observational unit is stored in multiple tables.
```{r}
students4 <- read.csv("students4.csv")
students4 # note that name, sec are repeated multiple times, so need to split this table into two: one for student info, and the other for grades
 studentinfo <- select(students4, id , name, sex)
 studentinfo <-unique(studentinfo)
 studentinfo
 
 gradebook <- select(students4, id, class, midterm, final)
 gradebook 
```

### Example from swirl() for cleaning up SAT data using chaining
```{r}
sat <- read.csv("sat.csv") #file is from the swirl() assignment and is in this repo
sat #note how the columns have mixed variables of part, sex, totals

# 1. select() all columns that do NOT contain the word "total" since we can get these from summing up once we tidy up the data
# 2. gather() all columns EXCEPT score_range, using key = part_sex and value = count.
# 3. separate() part_sex into two separate variables sat(columns), called "part" and "sex", respectively. You may need to check
# 4. Use group_by() (from dplyr) to group the data by part and sex, in that order.
# 5. Use mutate to add two new columns, whose values will be automatically computed group-by-group:
#   * total = sum(count)
#   * prop = count / total

sat %>%
  select(-contains("total")) %>%
  gather(part_sex, count, -score_range) %>%
  separate(part_sex, c("part", "sex")) %>%
  group_by(part, sex) %>%
  mutate(total = sum(count), prop = count / total) %>% 
  print
```







