---
title: "How to use sub, gsub, matches, and grepl"
author: "Arin Parsa"
date: "7/24/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## sub and gsub
sub() substitutes one occurrence of a pattern while gsub() substitutes all occurrences
```{r}
library(dplyr)

name <- "Arin:Parsa"
sub(":", " ", name)


name2 <- "George:Walker:Bush"
gsub(":", " ", name2)

sentence <- "the weather is pleasant."
gsub("^t", "T", sentence)

```

## matches and grepl
matches() and grepl() can be used to match regex patterns. Regex is a regular expression, which is a sequence of characters used in search patterns. It contains literals as well as metacharacters such as ^ for beginning of a sentence, $ for end of sentence, and many more.

matches() can be used in select() in dplyr to match regex patterns to select column variables while grepl() can be used anywhere including in filter() for rows
Below example shows how to use matches() and grepl()

```{r}
## Load GDP and educational datasets for countries
gdp <- read.csv("https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv")
edu <- read.csv("https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv")

gdp <- gdp[5:194,]

gdp <- gdp %>% rename(Rank = Gross.domestic.product.2012, CountryCode = X) %>% mutate(Rank = as.numeric(Rank))

## Merging the gdp and edu datasets
merge_df <- merge(gdp, edu, by = "CountryCode")

## Using matches() to select only the columns that have .Name in their titles
df_col_names <- merge_df %>% select(matches(".[Nn]ame"))
head(df_col_names, 5)

## Using grepl() to select the observations in the Special.Notes column that have the string "Fiscal year end: June" in them
merge_df1 <- merge_df %>% filter(grepl("Fiscal year end: June", Special.Notes))
merge_df1

```


