---
title: "Caching the inverse of a matrix"
author: "Arin Parsa"
date: "6/22/2020"
output: 
  html_document:
    toc: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment: Caching the Inverse of a Matrixless 
From JHU Data Science Course on Coursera

"Matrix inversion is usually a costly computation and there may be some benefit to caching the inverse of a matrix rather than compute it repeatedly (there are also alternatives to matrix inversion that we will not discuss here). Your assignment is to write a pair of functions that cache the inverse of a matrix.

Write the following functions:

* makeCacheMatrix: This function creates a special "matrix" object that can cache its inverse.
* cacheSolve: This function computes the inverse of the special "matrix" returned by * makeCacheMatrix above. If the inverse has already been calculated (and the matrix has not changed), then the cachesolve should retrieve the inverse from the cache."

```{r}
## makeCacheMatrix creates a special "matrix", which is a list
## containing a function to set and get the value of the matrix, and
## set and get the inverse of the matrix

makeCacheMatrix <- function(x = matrix()) {
  i <- NULL
  # set the matrix and clear the cache if any
  set <- function(y) {
    x <<- y    # Set the value
    i <<- NULL # Clear the cache
  }
  # get the matrix
  get <- function() x
  #set inverse in cache
  setInverse <- function(inverse) i <<- inverse
  # return cached inverse
  getInverse <- function() i
  
  #list of four functions to get and set matrix, get and set cached inverse
  list(set = set, get = get,
       setInverse = setInverse,
       getInverse = getInverse)
}

## cacheSolve calculates the inverse of the special "matrix"
## created with the makeCacheMatrix function. However, it first checks 
## to see if the inverse has already been calculated. If so, it gets 
## the inverse from the cache and skips the computation. Otherwise, 
## it calculates the inverse of the data (matrix) and sets the value of the inverse 
## in the cache via the setInverse function.

cacheSolve <- function(x, ...) {
  i <- x$getInverse() # This fetches the cached value for the inverse
  if(!is.null(i)) { # If the cache was not empty, we can just return it
    message("getting cached data")
    return(i)
  }
  # The cache was empty. We need to calculate it, cache it, and then return it.
  data <- x$get()  # Get value of matrix
  i <- solve(data) # Calculate inverse
  x$setInverse(i)  # Cache the result
  i    
}

```

## Calling cacheSolve function twice. Note that on second iteration, we get cached inverse

```{r}

specialMatrix <- makeCacheMatrix(matrix(c(1, 1, 1, 2), 2, 2))
cacheSolve(specialMatrix)
cacheSolve(specialMatrix)


```

