---
title: "Learning loop functions using Flags data set: lapply, sapply, tapply, vapply"
author: "Arin Parsa"
date: "6/6/2020"
output: 
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Flags data set by UCI ML Rep
http://archive.ics.uci.edu/ml/datasets/Flags

## Flags Data Set Information
This data file contains details of various nations and their flags. In this file the fields are separated by spaces (not commas). With this data you can try things like predicting the religion of a country from its size and the colors in its flag.

10 attributes are numeric-valued. The remainder are either Boolean- or nominal-valued.


Attribute Information:

1. name: Name of the country concerned
2. landmass: 1=N.America, 2=S.America, 3=Europe, 4=Africa, 4=Asia, 6=Oceania
3. zone: Geographic quadrant, based on Greenwich and the Equator; 1=NE, 2=SE, 3=SW, 4=NW
4. area: in thousands of square km
5. population: in round millions
6. language: 1=English, 2=Spanish, 3=French, 4=German, 5=Slavic, 6=Other Indo-European, 7=Chinese, 8=Arabic, 9=Japanese/Turkish/Finnish/Magyar, 10=Others
7. religion: 0=Catholic, 1=Other Christian, 2=Muslim, 3=Buddhist, 4=Hindu, 5=Ethnic, 6=Marxist, 7=Others
8. bars: Number of vertical bars in the flag
9. stripes: Number of horizontal stripes in the flag
10. colours: Number of different colours in the flag
11. red: 0 if red absent, 1 if red present in the flag
12. green: same for green
13. blue: same for blue
14. gold: same for gold (also yellow)
15. white: same for white
16. black: same for black
17. orange: same for orange (also brown)
18. mainhue: predominant colour in the flag (tie-breaks decided by taking the topmost hue, if that fails then the most central hue, and if that fails the leftmost hue)
19. circles: Number of circles in the flag
20. crosses: Number of (upright) crosses
21. saltires: Number of diagonal crosses
22. quarters: Number of quartered sections
23. sunstars: Number of sun or star symbols
24. crescent: 1 if a crescent moon symbol present, else 0
25. triangle: 1 if any triangles present, 0 otherwise
26. icon: 1 if an inanimate image present (e.g., a boat), otherwise 0
27. animate: 1 if an animate image (e.g., an eagle, a tree, a human hand) present, 0 otherwise
28. text: 1 if any letters or writing on the flag (e.g., a motto or slogan), 0 otherwise
29. topleft: colour in the top-left corner (moving right to decide tie-breaks)
30. botright: Colour in the bottom-left corner (moving left to decide tie-breaks)

## Create a vector of column names, and read file as table into a data frame
```{r}

colnames <- c("country", "landmass", "zone", "area", "population", "language", "religion", "bars", "stripes", "colors", "red", "green", "blue", "gold", "white", "black", "orange","mainhue", "circles", "crosses", "saltires", "quarters", "sunstars", "crescents", "triangle", "icon","animate", "text", "topleft", "botright")

# get flag data set from UCI. If this fails, get the dataset from the repo of this markdown file
dataset <-"https://archive.ics.uci.edu/ml/machine-learning-databases/flags/flag.data"

# read data into a table. 
flags <- read.table(dataset, sep=",", col.names = colnames, as.is = TRUE)

```
## Verify data is read properly and all countries are loaded
```{r}
flags[,1]
```

## Various operations to learn lapply and sapply
```{r}

# create a subset for data for flag colors
flag_colors <- flags[, 11:17]
head(flag_colors)

lapply(flag_colors, sum) #returns a list of number of flags based on color
sapply(flag_colors, sum) # sapply returns a vector since list value has only one element which the number of flags
sapply(flag_colors, mean) # find the mean of flag colors


flag_shapes <- flags[,19:23] #returns a data frame of flag shapes
lapply(flag_shapes, range)  #returns a list containing range of flag shapes (how many flags per shape)
shape_mat <-sapply(flag_shapes, range) # sapply on range will return a matrix this time since list element has two values (lower, upper values of range)
shape_mat 
unique(c(3, 4, 5, 5, 5, 6, 6)) #an example to show that unique function will return only 3, 4, 5, 6
unique_vals_list <- lapply(flags, unique) # returns a list of unique values per column in flags data frame
unique_vals_list

v <- sapply(unique_vals_list, length) #sapply returns a vector of number of unique values per column
v

l <- lapply(unique_vals_list, length) #lapply returns a list containing the number of unique values per column
l

v["landmass"] # returns number of unique landmasses. Note the use of [ since sapply(unique_vals_list, length) returned us a vector containing the number of unique elements per column


l[["landmass"]] # returns number of unique landmasses. Note the use of [[ since lapply(unique_vals_list, length)

sapply(flags,unique) #returns unique values in the flag data frame


lapply(unique_vals_list, function(elem) elem[2]) #using a custom function, return second element in the unique values

```
## Various examples to learn tapply
tapply allows us to run a function on a dataset and group by a subset
```{r}
populationbyreligion <- tapply(flags$population, flags$religion, sum) #returns sum of population by religion
populationbyreligion_and_lang <- tapply(flags$population, list(flags$religion, flags$language), sum) #groupby religion and language


```





